Data Descriptor: Corresponding author journal publishing activities 2014-2018

### 2019-10-27 analysis/paper.Rmd
add colleen's review

Assessing the volume and share of corresponding author publications is crucial for the planning of open access funding programs and licensing models [@Schimmer_2015]. However, funders and national library consortia often lack access to bibliometric data, which allow for determining the publication output per publisher, journal and journal business model across countries. Here, by obtaining country affiliations from the Web of Science at the level of journals, and combining them with normalized publisher and open access status information, we created a dataset about the productivity of corresponding authors for the period 2014-2018. This dataset is expected to have essential value for the quantitative understanding of scholarly publishing. It will not only help to analyze the market share of publishing houses, but will also uncover the current state of the transition of subscription-based publishing models to fully open access.

@@

There have also been analytical applications for such datasets. Major bibliometric databases including the Web of Science, Scopus and Dimension have started to integrate open access status information. In these databases, Unpaywall data is often used to identify openly available full-texts at the article-level, while the Directory of Open Access Journals provides information about if a journal is fully open access. In its most recent edition, the Leiden Ranking presents open access indicators at the level of universities [@leiden]. In addition to university-specific data, the German Open Access Monitor provides breakdowns by publisher and journal and includes publications from non-university research institutions based in Germany [@oa_monitor]. In its [Publisher Analytics Reports](https://clarivate.com/webofsciencegroup/solutions/publisher-analytics-reports/) Clarivate Analtics presents the performance of publishers across subjects, institutions and countries based on Web of Science data.

Here, we build an open dataset on the global productivity of corresponding authors. Drawing on Palzenberger's (2015) previous work, we not only created an updated dataset about the number and proportion of scholarly articles per country of affiliation from corresponding authors. The dataset also allows for a breakdown by journal, including information whether it was operated as a subscription or fully open access journal, and by publishers. We constructed the dataset using the quality-assured Web of Science in-house database from the German Competence Center for Bibliometrics. We enriched the data with publisher information from Crossref, a large DOI registration agency for scholarly work, and with open access journal information provided by the ISSN-GOLD-OA list, a dataset that merges open access journal information from the DOAJ, OpenAPC, PubMed and ROAD [@issn].

The data does not only provide background information for negotiations in the context of transformative agreements, but can also be used in the quantitative study of the scholarly publishing industry and the open access uptake across countries. For instance, national transformative agreements with large publishers will increase opportunities to publish open access. The existence of such agreements will therefore likely have an impact on the open access share of a country with such agreements. To demonstrate the dataset's significance, we will present two new findings using the dataset: a global map of publishers' market shares, and an analysis of country of affiliation by publisher. The source code used for compiling the dataset is shared along with the resulting datasets.

@@

Like @Palzenberger_2015, the above-listed selection reflects data needs in the context of open access negotiations between library consortia and publishers. The restriction to the article types original article and review is in accordance with most open access contracts where other types of journal content like letters or meeting abstracts are not considered (see for instance @Sander_2019).

To allocate open access costs to institutions, @Schimmer_2015 proposed to use affiliations from corresponding authors, presuming that corresponding author affiliations can be counted straight. Under this counting method, only one institution or one country receives full credit for a publication [@Huang_2011]. However, there have been some concerns with using corresponding authorship’s data from bibliometric databases for cost analysis recently. @Gumpenberger_2018 noted two potential issues: multiple corresponding author affiliations and multiple corresponding authors per article.

Indeed, the Web of Science has begun to systematically keep track of more than one corresponding author recently: We found that the Web of Science did record more than one corresponding author for 6.2% of indexed articles published between 2014 - 2018. However, Figure 1 suggests that indexing of more than one corresponding author started in 2016. Moreover, a 5% of corresponding authors author was internationally co-located (see Figure 2). We therefore decided to use whole counting where every original article or review was counted once per country of affiliation of the corresponding author(s).

@@

## Data and Code availability

The underlying source code in R and SQL for both testing the retrieval and compiling the datasets are openly shared as executable reports, making our data work more reproducible and transparent. The data analytics work, which was organized as research compendium [@Marwick_2017] using the holepunch R package [@holepunch], is accessible at <https://github.com/subugoe/oa2020cadata>.

@@

1. Initial exploration about how corresponding authorships are represented in the WoS-KB ([`001_kb_rp_pretest.Rmd`](analysis/001_kb_rp_pretest.Rmd))
2. Coverage analysis of reprint authors in the WoS-KB for the period 2014 - 2018 ([`002_kb_rp_coverage.Rmd`](analysis/002_kb_rp_coverage.Rmd)).
3. Steps to obtain and compile the dataset about corresponding author country affiliations 2014 - 2018 ([`003_kb_fetch_ca.Rmd`](analysis/003_kb_fetch_ca.Rmd))
4. Steps to obtain and compile the dataset about the global publisher output 2014 - 2018 ([`004_kb_fetch_publisher.Rmd`](analysis/004_kb_fetch_publisher.Rmd)).

@@

The first file, named `journal_publisher_14_18`, contains information about the publishing market landscape for the period 2014-2018 in terms of the aggregated number of articles and reviews  by year and journal represented by the ISSN identifier indexed in the Web of Science. Table 1 describes the data variables in detail. We were able to retrieve publisher and journal title from Crossref for `r publisher_df %>% filter(!is.na(publisher)) %>% distinct(issn_l) %>% nrow() / publisher_df %>% distinct(issn_l) %>% nrow() * 100` %
 of all journals, representing `r publisher_df %>% filter(!is.na(publisher)) %>% .$articles %>% sum() / sum(publisher_df$articles) * 100` % of the overall article volume. `r publisher_df %>% filter(oa_journal == TRUE) %>% distinct(issn_l) %>% nrow()` out of `r publisher_df %>% filter(!is.na(publisher)) %>% distinct(issn_l) %>% nrow()`
 investigated journals with `ISSN-L` were identified as fully open access journals. In total, the dataset provides information for `r sum(publisher_df$articles)` records indexed in the Web of Science with journal information in the five-years period 2014-2018..
 <!-- are only journals with valid issn considered? Are there journals in the dataset without issn-l?-->
<!--what about the 9% that could not be matched with crossref? are they excluded or included with WoS information /without journal information?-- per Wos ISSN-->
Data Schema `journal_publisher_14_18`:

@@

The dataset named `rp_publisher_14_18` provides a breakdown of journal information aggregated by  countries of affiliation. Only corresponding authorships, indicated by the role "reprint author" in the Web of Science database, were taken into consideration. The following table  describes the variables in detail. Overall, 99 % of records representing original articles and reviews in the period 2014 - 2018 had affiliation information for corresponding authors at the country-level.

@@

Based on the dataset `journal_publisher_14_18` the global market share of publishing houses in terms of articles and reviews indexed in the Web of Science can be analyzed for the period 2014-18. The following table  shows the overall market share for the ten largest publishers. The following figure presents a breakdown by journal business model, differentiating between subscription journals and fully open access journals.

```{r, results='asis'}
my_df <- readr::read_csv("../data/journal_publisher_14_18.csv")
tt_all <- my_df %>%
  group_by(publisher) %>%
  summarise(n = sum(articles)) %>%
  mutate(prop = n / sum(n))  %>%
  mutate(publisher = ifelse(is.na(publisher), "Other", publisher)) %>%
  arrange(desc(n)) %>%
  mutate(publisher_group = fct_other(publisher, keep = publisher[1:11])) %>%
  ungroup() %>%
  group_by(publisher_group) %>%
  summarise(n = sum(n), prop = sum(prop)) %>%
  mutate(publisher_group = fct_reorder(publisher_group, n, )) %>%
  mutate(publisher_group = fct_relevel(publisher_group, "Other")) %>%
  arrange(desc(publisher_group))
tt_all %>%
  mutate(prop = prop * 100) %>%
  janitor::adorn_totals() %>%
  knitr::kable(
    caption = "Publisher Market Share 2014 - 2018 in terms of articles and reviews indexed in the Web of Science",
    col.names = c("Publisher", "Articles and Reviews published", "Percentage"),
    format.args = list(big.mark = ",")
  )
```

@@

Our data confirms that the global publishing landscape is dominated by a few large publishers. In total, ten publishers accounted for around two-thirds of articles published between 2014 - 2018. The Big Three, Elsevier, Springer Nature and Wiley, secured a market share of 46 %. However, further analysis suggests large differences between subscription and fully open access journals: The market for fully open access journals seems to be less dominated by major publishing houses. A notable exception is Springer Nature with its large fully open access portfolio including BioMed Central and the large mega-journals Scientific Reports and Nature Communications.

@@

Using `rp_publisher_14_18.csv` the publisher market shares can be further broken down by country affiliation from corresponding authors. The following figure presents a global map of scholarly journal publishing represented in the Web of Science. Publications from institutions based in the European Union were summarized as "EU28". Together, all EU 28 member states constituted the largest market for journal publishing between 2014-18.

@@

The following figure shows the country share per publisher in terms of original articles and reviews published in subscription journals using corresponding author affiliations. The flipped x-axis shows the top 10 countries of affiliations in terms of articles published, while the y-axis depicts the percentage per publisher. The figure indicates, which countries were particularly important for a publisher. Furthermore, it suggests large variations across publishers. Take for instance the publisher Royal Society of Chemistry. Around every third article published in a subscription journal was submitted from an author based in China, while the overall share of China-based authors was significantly lower with `r rp_df %>% filter(oa_journal == FALSE) %>% group_by(country_code) %>% summarise(n = sum(n_publications)) %>% arrange(desc(n)) %>% mutate(prop = n /sum(n)) %>% filter(country_code == "CHN") %>% .$prop * 100` %.

@@

@@

Although the Web of Science database is a well-established bibliometric database it has certain limitations that need careful consideration when using and interpreting our data. First, the Web of Science is selective in terms of journal inclusion and is not covering the whole journal portfolio of the investigated publishers. There is also evidence about a considerable indexing lag of non-English journal publications relevant to domestic research [@Lopez_Pineiro_2014]. Although publisher names were disambiguated, journal title transfers between publishers were not tracked. Furthermore, the year of inclusion in a journal issue, i.e. print publication date, was used. However, there can be a considerable time-lag between online-first publication and issue inclusion.

Moreover, as is the case with any other bibliometric database, the data structure can be subject to change. In particular, our pretests suggest that the Web of Science author role field used to identify corresponding authorships has undergone a change recently, starting to cover more than one reprint or corresponding author per article in 2016.

The ISSN GOLD-OA list [@issn] was used to determine fully open access journals. However, it does not contain information when a subscription-based journal became fully open access. The figures should therefore be interpreted as numbers of articles available in current fully open access journals, instead of published in a fully open access journal at the time of publication. Open access article counts are therefore upper estimates. <!-- Is it really upper estimates? What about reverse flip?-->

@@

>>>>

Assessing the volume and share of corresponding author publications is crucial for the planning of open access funding programs and licensing models [@Schimmer_2015]. However, funders and national library consortia often lack access to bibliometric data, which allow for determining the publication output per publisher, journal and journal business model across countries. Here, by obtaining country affiliations from the Web of Science, and combining them with normalized publisher and open access status information for journals, we created a dataset about the productivity of corresponding authors for the period 2014-2018. This dataset is expected to have essential value for the quantitative understanding of scholarly publishing. It will not only help to analyze the market share of publishing houses, but will also uncover the current state of the transition of subscription-based publishing models to fully open access.

@@

There have also been analytical applications for such datasets. Major bibliometric databases including the Web of Science, Scopus and Dimension have started to integrate open access status information. In these databases, Unpaywall data is often used to identify openly available full-texts at the article-level, while the Directory of Open Access Journals provides information on whether a journal is fully open access. In its most recent edition, the Leiden Ranking presents open access indicators at the level of universities [@leiden]. In addition to university-specific data, the German Open Access Monitor provides breakdowns by publisher and journal and includes publications from non-university research institutions based in Germany [@oa_monitor]. In its [Publisher Analytics Reports](https://clarivate.com/webofsciencegroup/solutions/publisher-analytics-reports/) Clarivate Analtics presents the performance of publishers across subjects, institutions and countries based on Web of Science data.

Here, we build an open dataset on the global productivity of corresponding authors. Drawing on Palzenberger's (2015) previous work, we created an updated dataset that reveals the number and proportion of scholarly articles published per country, according to the corresponding author affiliation. The dataset also allows for a global breakdown by journal, including information whether it was operated as a subscription or fully open access journal, and by publishers. We constructed the dataset using the quality-assured Web of Science in-house database from the German Competence Center for Bibliometrics. We enriched the data with publisher information from Crossref, a large DOI registration agency for scholarly work, and with open access journal information provided by the ISSN-GOLD-OA list, a dataset that merges open access journal information from the DOAJ, OpenAPC, PubMed and ROAD [@issn].

The data not only provides background information for negotiations in the context of transformative agreements, but can also be used in the quantitative study of the scholarly publishing industry and the uptake of open access publishing across countries. For instance, national transformative agreements with large publishers will increase authors'  opportunities to publish open access, and the existence of such agreements is therefore likely to have an impact on the open access share of a country with such agreements. To demonstrate the dataset's significance, we will present two new findings using the dataset: a global map of publishers' market shares, and an analysis of country of affiliation by publisher. The source code used for compiling the dataset is shared along with the resulting datasets.

@@

Like @Palzenberger_2015, the above-listed selection reflects data needs in the context of open access negotiations between library consortia and publishers. The restriction to the article types original article and review is in accordance with most open access contracts where other types of journal content, such as letters or meeting abstracts, are not considered (see for instance @Sander_2019).

To allocate open access costs to institutions, @Schimmer_2015 proposed using affiliations from corresponding authors, presuming that corresponding author affiliations can be counted straight. Under this counting method, only one institution or one country receives full credit for a publication [@Huang_2011]. However, there have been some concerns raised over using corresponding authorship’s data from bibliometric databases for cost analysis recently. @Gumpenberger_2018 noted two potential issues: multiple corresponding author affiliations and multiple corresponding authors per article.

Indeed, the Web of Science has recently begun to systematically keep track of more than one corresponding author per paper: We have found that the Web of Science recorded more than one corresponding author for 6.2% of indexed articles published between 2014 - 2018. However, Figure 1 suggests that indexing of more than one corresponding author started in 2016. Moreover, 5% of corresponding authors had multiple affiliations from different countries (see Figure 2). We therefore decided to use whole counting in which every original article or review was counted once per country of affiliation of the corresponding author(s).

@@

## Data and code availability

The underlying source code in R and SQL for both testing the retrieval and compiling the datasets are openly shared as executable reports, making our data work more reproducible and transparent. The data analytics work, which was organized as a research compendium [@Marwick_2017] using the holepunch R package [@holepunch], is accessible at <https://github.com/subugoe/oa2020cadata>.

@@

1. Initial exploration on how corresponding authorships are represented in the WoS-KB ([`001_kb_rp_pretest.Rmd`](analysis/001_kb_rp_pretest.Rmd))
2. Coverage analysis of reprint authors in the WoS-KB for the period 2014 - 2018 ([`002_kb_rp_coverage.Rmd`](analysis/002_kb_rp_coverage.Rmd)).
3. Steps to obtain and compile the dataset relative to corresponding author country affiliations 2014 - 2018 ([`003_kb_fetch_ca.Rmd`](analysis/003_kb_fetch_ca.Rmd))
4. Steps to obtain and compile the dataset relative to the global publisher output 2014 - 2018 ([`004_kb_fetch_publisher.Rmd`](analysis/004_kb_fetch_publisher.Rmd)).

@@

The first file, named `journal_publisher_14_18`, contains information about the publishing market landscape for the period 2014-2018 in terms of the aggregated number of articles and reviews  by year and journal represented by the ISSN identifier indexed in the Web of Science. Table 1 describes the data variables in detail. We were able to retrieve publisher and journal title from Crossref for `r publisher_df %>% filter(!is.na(publisher)) %>% distinct(issn_l) %>% nrow() / publisher_df %>% distinct(issn_l) %>% nrow() * 100`%
 of all journals, representing `r publisher_df %>% filter(!is.na(publisher)) %>% .$articles %>% sum() / sum(publisher_df$articles) * 100` % of the overall article volume. `r publisher_df %>% filter(oa_journal == TRUE) %>% distinct(issn_l) %>% nrow()` out of `r publisher_df %>% filter(!is.na(publisher)) %>% distinct(issn_l) %>% nrow()`
 investigated journals with `ISSN-L` were identified as fully open access journals. In total, the dataset provides information for `r sum(publisher_df$articles)` records indexed in the Web of Science with journal information in the five-year period 2014-2018. Journals that could not be matched with the `ISSN-L` or Crossref are marked with `NA`.

<!-- are only journals with valid issn considered? Are there journals in the dataset without issn-l?-->
<!--what about the 9% that could not be matched with crossref? are they excluded or included with WoS information /without journal information?-- per Wos ISSN-->
Data Schema `journal_publisher_14_18`:

@@

The dataset named `rp_publisher_14_18` provides a breakdown of journal information aggregated by  countries of affiliation. Only corresponding authorships, indicated by the role "reprint author" in the Web of Science database, were taken into consideration. The following table  describes the variables in detail. Overall, 99% of records representing original articles and reviews in the period 2014 - 2018 had affiliation information for corresponding authors at the country-level.

@@

Based on the dataset `journal_publisher_14_18` the global market share of publishing houses as measured by articles and reviews indexed in the Web of Science can be analyzed for the period 2014-18. The following table shows the overall market share of the ten largest publishers. The following figure presents a breakdown by journal business model, differentiating between subscription journals and fully open access journals.

```{r, results='asis'}
my_df <- readr::read_csv("../data/journal_publisher_14_18.csv")
tt_all <- my_df %>%
  group_by(publisher) %>%
  summarise(n = sum(articles)) %>%
  mutate(prop = n / sum(n))  %>%
  mutate(publisher = ifelse(is.na(publisher), "Other", publisher)) %>%
  arrange(desc(n)) %>%
  mutate(publisher_group = fct_other(publisher, keep = publisher[1:11])) %>%
  ungroup() %>%
  group_by(publisher_group) %>%
  summarise(n = sum(n), prop = sum(prop)) %>%
  mutate(publisher_group = fct_reorder(publisher_group, n, )) %>%
  mutate(publisher_group = fct_relevel(publisher_group, "Other")) %>%
  arrange(desc(publisher_group))
tt_all %>%
  mutate(prop = prop * 100) %>%
  janitor::adorn_totals() %>%
  knitr::kable(
    caption = "Publisher Market Share 2014 - 2018 as measured by articles and reviews indexed in the Web of Science",
    col.names = c("Publisher", "Articles and Reviews published", "Percentage"),
    format.args = list(big.mark = ",")
  )
```

@@

Our data confirms that the global publishing landscape is dominated by a few large publishers. In total, ten publishers accounted for around two-thirds of articles published between 2014 - 2018. The Big Three, Elsevier, Springer Nature and Wiley, secured a market share of 46 %. However, further analysis suggests large differences between subscription and fully open access journals: the market of fully open access journals seems to be less dominated by major publishing houses. A notable exception is Springer Nature with its large fully open access portfolio including BioMed Central and the large mega-journals Scientific Reports and Nature Communications.

@@

Using `rp_publisher_14_18.csv` the publisher market shares can be further broken down by country affiliation of corresponding authors. The following figure presents a global map of scholarly journal publishing represented in the Web of Science. Publications from institutions based in the European Union were summarized as "EU28", and, together, constitute the largest market for journal publishing between 2014-18. The figure further highlights the consistency of market dominance of the same set of large publishers across all regions of the world.

@@

The following figure shows the country share per publisher as measured by original articles and reviews published in subscription journals using corresponding author affiliations. The flipped x-axis shows the top 10 based on affiliations and articles published, while the y-axis depicts the percentage per publisher. The figure indicates, which countries were of particular relevance for a publisher. Furthermore, it suggests large variations across publishers. Take, for instance, the publisher Royal Society of Chemistry (RSC). Around every third article published in a subscription journal of RSC was submitted by an author based in China, while the overall share of China-based authors was significantly lower with `r rp_df %>% filter(oa_journal == FALSE) %>% group_by(country_code) %>% summarise(n = sum(n_publications)) %>% arrange(desc(n)) %>% mutate(prop = n /sum(n)) %>% filter(country_code == "CHN") %>% .$prop * 100` %.

@@

### Discussion and outlook

As country-level consortia increasingly embrace transformative agreements as a means to increase the proportion of their research published open access [^1], the dataset constitutes a highly useful resource for researchers and practitioners alike. When combined with information on a country’s own subscription and hybrid APC expenditure, bibliometric data about corresponding author affiliation benefit cost modelling for a fully open access future national consortia can derive key insights such as

- What is the total research output of our authors?
- How fast is our output growing?
- What percentage of our authorship is as corresponding author (responsible for open access publishing costs)?
- How much of our output is already published fully open access?
- With which publishers do our authors most frequently publish?
- How does the publisher distribution of our output relate to our subscription expenditure?
- Would a shift to open access with publisher X cost more, less or the same as what we currently spend on subscriptions (and hybrid APCs)?
- What value can we place on authorship and open access publishing services?

@@

Although the Web of Science database is a well-established bibliometric database it has certain limitations that need careful consideration when using and interpreting our data. First, the Web of Science is selective in terms of journal inclusion and does not covering the whole journal portfolio of the investigated publishers. There is also evidence about a considerable indexing gap of non-English journal publications relevant to domestic research [@Lopez_Pineiro_2014]. Although publisher names were disambiguated, journal title transfers between publishers were not tracked. Furthermore, the year of inclusion in a journal issue, i.e. print publication date, was used. However, there can be a considerable time-lag between online-first publication and issue inclusion.

Moreover, as is the case with any other bibliometric database, the data structure can be subject to change. In particular, as mentioned above, our pretests suggest that the Web of Science author role field used to identify corresponding authorships has undergone a change recently, starting to cover more than one reprint or corresponding author per article in 2016.

The ISSN GOLD-OA list [@issn] was used to determine fully open access journals. However, it does not contain information when a subscription-based journal became fully open access. The figures should therefore be interpreted as numbers of articles available in current fully open access journals, rather than as published in a fully open access journal at the time of publication. <!-- Open access article counts are therefore upper estimates.  Is it really upper estimates? What about reverse flip?-->

@@

[^1]: See the ESAC registry for an up-to-date list of transformative agreements from national consortia <https://esac-initiative.org/about/transformative-agreements/agreement-registry/>
